Classification and characterization of anthropogenic plant communities in the northwestern Iberian Peninsula

### 2024-11-15 doc/response.Rmd

---
title: "Response letter on the manuscript Classification and characterization of anthropogenic plant communities at the ecoregion level (2nd revision)"
output: 
  bookdown::pdf_document2:
    number_sections: false
    toc: false
bibliography: refs.bib
csl: journal-of-vegetation-science.csl
link-citations: yes
always_allow_html: true
---

@@

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r message = FALSE, echo = FALSE, warning = FALSE}
knitr::knit_hooks$set(inline = function(x) {
  prettyNum(x, big.mark = ",")
})
```

# Co-ordinating Editor Comments to Author:

*Dear authors, your work has improved consistently. However, in this round, I involved one additional referee, Zdenka Lososová, who pointed attention to some critical points of discussion. Please revise the manuscript according to her requests.*

\textcolor{teal}{Thank you for a new chance to improve our manuscript. We have implemented all the changes required by the new reviewer, as well as the change requested by reviewer 2 regarding the synoptic table.}

# Referee: 1

*I consider the paper has been improved and the flaws were seen in the first review have been corrected*

\textcolor{teal}{Thank you again for your contribution to improving our manuscript.}

# Referee: 3

*The manuscript provides a detailed overview of anthropogenic vegetation types in the northwestern Iberian Peninsula. The authors assembled a robust dataset of plots. These plots were classified using semi-supervised classification methods and assigned to 25 alliances. The primary characteristics of these alliances are given using proportions of species with different origins, traits, and indicator values. I appreciate the way the classification procedure is described. I also appreciate the nice form of all of the figures. I read the manuscript with all the supplements and the response letter, and I see that the authors highly improved the manuscript. I did not review the manuscript in the first round. However, I have some serious comments which should be discussed.*

\textcolor{teal}{Thank you for your kind words and your advice to improve the manuscript.}

*The class Galio-Urticetea is not included as it also includes native communities. However, the same is true for Bidentetea, and this class is included. Explain!*

\textcolor{teal}{The class Galio-Urticetea is included in the manuscript. However, we are following the vegetation classification by Mucina et al. (https://doi.org/10.1111/avsc.12257) which includes Galio-Urticetea within the class Epilobietea angustifolii. We have modified the text (page 7, line 11-12) to make this clear.}

*Hemicryptophytes are the most numerous group of species in the dataset. This fact is repeatedly mentioned in the text of results and discussion. However, hemicryptophytes are not shown in Fig. 4. It is confusing and not logical. I would strongly suggest redrawing the figure and replacing the geophytes with hemicryptophytes (or merging them into one category of perennial herbs). / p. 12, l. 20. I mentioned above, this is not good decision. It is shown that the hemicryptophytes are the most common life form!*

\textcolor{teal}{We have done as requested, and included hemicryptophytes in Figure 4. We have kept geophytes as a separate category in Figure 4 because they are an important part of the definition of the alliance Allion triquetri.}

*A flow diagram (or table) showing the translation between alliances obtained by different classification schemes and EUNIS types would be helpful for the readers.*

\textcolor{teal}{The translation between alliances and EUNIS types is provided in Appendix S3, we have revised the text (page 13, line 8) to make this clearer.}

*I am not comfortable with the use of Central European historical data for the Iberian peninsula, as described in the introduction. A lot of Central European weed species originate in South Europe, not only in the Near East, and the spread of archaeophytes was here probably earlier. It should be at least mentioned. (p. 3, l. 9-16) / p. 12, l. 12 Chytrý et al. 2021? Is it an appropriate source for the Iberian peninsula? I doubt it. The delimitation of alien species in CZ could hardly work for other regions. Moreover, the authors of Catalogue of alien species in CZ is Pyšek et al. – the newest version was published in Preslia in 2022.*

\textcolor{teal}{Applying the concept of archaeophyte in southern Europe is problematic. However, we wanted to consider this category, solely for the purpose of comparison with similar studies in other European regions. Unfortunately, historical vegetation data is not as available in the Iberian Atlantic territories as it is in other parts of Europe, and this is the reason why we used the catalogues for the Czech Republic and Britain. We must clarify that we did not take those catalogues at face value. We used them to create a starting list of archaeophyte candidates. Then, we revised this list using Flora iberica, to remove species that are considered native in the Mediterranean part of the Iberian Peninsula. We classified the remaining species as putative archaeophytes in the study region, and only for the purposes of comparison. We have revised the text (page 11 line 19 to page 12 line 5) to better explain this, and to further stress that the classification must be interpreted with care (page 12, line 5-10). We have also updated the reference for the Czech catalogue.}

*Neophytes could originate from all the continents, not only from America, Africa, and Asia (p. 3, l. 19)*

\textcolor{teal}{We have removed this sentence.}

*P5, l. 18, 19 the format of references is not in appropriate format*

\textcolor{teal}{We are using a reference manager with AVS's official style template. After some time trying, we have not been able to fix these specific references. We will fix them manually if the manuscript is accepted.}

*p. 6, l. 11-12 the mix of two different taxonomic concepts is not the best option*

\textcolor{teal}{We agree in principle, however, in this case the taxa revised with POWO were a few non-European introduced taxa, therefore, this has not affected in any way the classification and the results of the article.}

*p. 10, l. 9-16. this part of the text belongs to the results and not to the methodological part*

\textcolor{teal}{We have moved the text to results as requested (page 13 line 10-17).}

*p. 14, l. 8. It would be helpful to explain in the methods which taxa are those problematic ones that are merged to some aggregates.*

\textcolor{teal}{Thank you for spotting this. The vegetation dataset that we use in this manuscript is part of a wider vegetation dataset for the whole Iberian Atlantic region (including all vegetation types, not just synanthropic). When we cleaned and curated the whole dataset, we did create some taxa aggregates. However, we have checked the plots and the taxa included in this manuscript and, for this subset, each aggregate can only correspond to one plant species. Therefore, all mentions to "taxa aggregates" can be changed to "species". We have done so in the text.}

*Do all of the plots include bryophytes? It is not described in the methods. Bryophytes are highly rare in the annual vegetation of disturbed sites.*

\textcolor{teal}{The SIVIM database does not provide consistent information on whether bryophytes had been recorded or not, but it is very likely that they were not in many cases. We have revised the manuscript to include this information (page 16, line 3). Bryophytes are rare in most synanthropic vegetation types, but they were relatively frequent and abundant in the vegetation of mesic trampled sites, i.e. Saginion procumbentis with Bryum argenteum.}

*p. 14, l. 12-13 The comparison of species numbers per plot of different sizes is inappropriate. The authors minimize the variance of the plot sizes, but still, could we be sure that the low number of species is not a result of lower plot sizes?*

\textcolor{teal}{We have revised the calculation to use only those plots that had the same and most frequent size value (= 20 m2) and recalculated the numbers of species accordingly.}

*References. Please fix the correct form of some authors' names in the references. / Němec, R., Lososová, Z., Dřevojan, P & Žáková, K. / Šilc U. / Šilc et al. – check all the names / Simonová D. / Tabasevic et al. – check / Zaliberova – check.*

\textcolor{teal}{Thank you very much for revising the spelling of the names. We have corrected them.}

*Appendix S5. Please add a header to the table with a short description and explanation of what the numbers in the table show. Define the phi value for diagnostic taxa and explain the meaning of the different colors used. Please note that the phi value is a number between 0 and 1. The table probably does not include the complete list of species. Some explanation is needed.*

\textcolor{teal}{We have added a header explaining all the requested points.}

# Referee: 2

*I couldn't find a Response Letter. If the authors haven't submitted one, I would ask them to do so. However, from the track-changes file I see that the authors addressed most of my previous points. The only remaining issue is the synoptic table: The authors included such a table as Appendix S5, but it only shows the phi values. In my opinion, a synoptic table should show the percentage frequency, and phi values reaching a certain threshold should be indicated by shading.*

\textcolor{teal}{We are sorry that the reviewer couldn't access the response letter. We have revised the synoptic table following the reviewer's specifications.}

### 2024-11-15 doc/manuscript.Rmd
AVS R2

* *Results* We assigned 2,086 vegetation plots to 25 anthropogenic alliances representing 9 vegetation classes (*Cymbalario-Parietarietea diffusae*, *Polygono-Poetea annuae*, *Papaveretea rhoeadis*, *Digitario sanguinalis-Eragrostietea minoris*, *Chenopodietea*, *Sisymbrietea*, *Bidentetea*, *Artemisietea vulgaris* and *Epilobietea angustifolii*). The plots included 1,149 taxa: 78% natives, 15% archaeophytes and 7% neophytes. Vegetation groups were organized along a principal axis of abiotic stress (dry-sunny to moist-shady habitats) and a secondary axis of disturbance.

@@

In temperate Europe, anthropogenic vegetation is a melting pot of floras from different biogeographical origins. The starting native pool of species favored by humans has been constantly enriched by human-assisted arrival of alien (i.e. non-native) species [@RN5504]. The introduction of archaeophytes (i.e. alien species arriving before 1.500 CE) likely started in the early Neolithic and peaked in the Bronze Age, with successive waves during periods of accelerated human colonization such as the early Middle Ages [@RN5504]. In Europe, synanthropic archaeophytes include many Mediterranean taxa originating from the segetal flora of the Near East [@RN5494]. Some of these archaeophytes are 'obligate weeds' that do not occur in non-anthropogenic habitats and are assumed to have evolved together with human agriculture [@RN5494]. The synanthropic flora of Europe has been further enriched by neophytes, alien taxa that arrived after the expansion of intercontinental trade in the Modern Age (after 1.500 CE) [@RN5439]. Synanthropic neophytes therefore include many American, African and Asian taxa that arrived in Europe because of the global trade networks established by western European colonialism [@RN5451]. In the last century, the European anthropogenic vegetation has shown a decrease in species richness and diversity [@RN5423], as rare synanthropic species have become rarer and neophytes have increased their abundance at the expense of natives and archaeophytes [@RN5441].

@@

To circumscribe our study vegetation, we followed the definition of anthropogenic vegetation in the revised classification of the vegetation of Europe by Mucina *et al.* [-@RN4698]. In our study area, this potentially includes the vegetation classes *Polygono-Poetea annuae*, *Papaveretea rhoeadis*, *Digitario sanguinalis-Eragrostietea minoris*, *Chenopodietea*, *Sisymbrietea*, *Bidentetea*, *Artemisietea vulgaris* and *Epilobietea angustifolii*. For the sake of completeness, we also considered the class *Cymbalario-Parietarietea diffusae* included by Mucina *et al.* [-@RN4698] in the vegetation of rock crevices and screes, since this class encompasses the vegetation of human-made walls.

@@

The main aim of our classification was for it to be consistent with regional phytosociological expertise and to match that expertise with current EuroVegChecklist alliances [@RN4698]. To achieve this, one advantage of using the SIVIM database is that a majority of the plots had been assigned to associations or alliances by the original authors [@RN5374]. However, there could be misclassification due to unclear concepts or human error. Therefore, we used TWINSPAN to revise the classification by the original authors, to define the number of clusters that better aligned with our syntaxonomical checklist and to remove outliers or misclassified plots.

We started this stage of the analyses by keeping only those plots (n = 2,095) that had been assigned by the original authors of the plot to any of the 9 anthropogenic vegetation classes defined in our syntaxonomical checklist (i.e., we removed the plots without syntaxa, that had been retrieved solely by the expert system). Then, we performed a TWINSPAN classification (with the same settings as above) within each vegetation class, attempting to classify its plots into a number of clusters equal to the number of alliances of that class in our checklist. We assigned the resulting clusters to an alliance in our checklist based on their composition of diagnostic species as described in the regional literature. In some cases, this resulted in a classification into a number of clusters in which (1) more than one cluster was assigned to the same alliance in the checklist; and (2) some alliances in the checklist could not be assigned to any of the clusters (i.e. the plots that had been assigned by the original authors to that alliance were separated in different clusters). By the end of this process, of the 34 alliances in our checklist, 25 had been assigned to clusters. Nine alliances were not assigned to any cluster. Of these, eight had been assigned to a few plots by the original authors: *Onopordion castellani* (n = 6 plots), *Onopordion acanthii* (5), *Euphorbion prostratae* (4), *Caucalidion lappulae* (4), *Roemerion hybridae* (2), *Hordeion murini* (2), *Taeniathero-Aegilopion geniculatae* (1) and *Fragarion vescae* (1). The alliance *Aegopodion podagrariae* had been assigned by the original authors to 39 plots, but our TWINSPAN analysis could not separate this alliance from the alliances *Geo urbani-Alliarion officinalis* and *Balloto-Conion maculati*.

This stage of the analyses allowed us to identify those plots in which there was an agreement between (1) our TWINSPAN-based classification (i.e. numerical classification) and (2) the classification based on the original author's syntaxon (i.e. regional phytosociological expertise). These plots (n = 1,850) were the best representatives of each alliance and became our training dataset for semi-supervised classification.

@@

Next, we conducted a semi-supervised classification [@RN5593] of the whole dataset (n = 2,508 plots) into the 25 TWINSPAN-validated anthropogenic alliances. Semi-supervised classification uses a training subset of *a priori* classified vegetation plots to classify a secondary subset of unclassified plots [@RN5383]. Since our goal was to refine the classification of the whole dataset, we allowed plots from the training subset to be re-assigned to other alliances during the re-classification. In addition, semi-supervised classification can create new groups to place data points that do not match the already existing *a priori* groups, but attempts to do so resulted in new groups with no ecological significance and thus we kept the 25 alliances as final anthropogenic vegetation groups. Furthermore, we applied a noise clustering fuzzy algorithm, which allows plots to be classified either into the *a priori* alliances or into a *noise* group which includes outliers and transitional plots [@RN5382]. We set the fuzziness coefficient to a low value (m = 1) to accommodate a high number of transitional plots; and we set the distance to the noise class to d = 1. We performed this analysis with the R package *vegclust* [@RN5383]. The semi-supervised classification with noise clustering resulted in the final classification of 2,086 plots into 26 alliances and 422 plots left out in the noise group. We used pairwise PERMANOVA (with 100,000 iterations, Euclidean distances and Holm's p-value correction) fitted with the R package *RVAideMemoire* [@RN5381] to test the significance of the final vegetation alliances; along with Principal Component Analysis (PCA) as implemented in the R package *FactoMineR* [@RN3166] to visualize the relationships between the classes and alliances.

@@

We classified the species as native, archaeophytes or neophytes in the Iberian Atlantic territories using the information in *Flora iberica* [-@RN3243] and catalogues of archaeophytes for Britain [@RN5490] and the Czech Republic [@RN5495]. We must stress that identifying archaeophytes in southern Europe is highly problematic [@RN5444] and our classification must be taken as an indication of putative archaeophyte character for the purposes of vegetation description and comparison with other European regions, rather than a definitive classification of the species; as stated by Preston *et al.* [@RN5490], the classification of a species as an archaeophyte should be interpreted as a hypothesis to be tested by further studies.

## Species life form

Using *FloraVeg.EU* [-@RN5385] we extracted the species’ life forms and kept *therophyte* and *geophyte* for further analysis since these two categories had the largest contribution to variance among alliances, as per an exploratory Principal Component Analysis (PCA) performed using the R package *FactoMineR* [@RN3166].

## Species height and flowering phenology

From *Flora iberica* [-@RN3243] we extracted the maximum height, median month of flowering and length of the flowering period of each species in the dataset. We used these variables to calculate plot-level community-weighted means (weighting by species cover) for each trait, to characterize the vegetation height and flowering phenology of the anthropogenic plant communities.

@@

The semi-supervised classification resulted in the classification of 2,086 vegetation plots into 25 anthropogenic alliances (**Fig. 1**, description of the alliances and correspondence with EUNIS habitat types in **Appendix S3**, groups of characteristic species in **Appendix S4**, synoptic table in **Appendix S5**) representing 9 vegetation classes. The regional literature included in our dataset recognized 69 associations (**Appendix S6**) within these alliances, but exploratory analysis (not shown) indicated that the separation of these associations within the alliances had a generally weak support based on numerical classification methods.

The classified plots included 1,149 taxa or taxa aggregates. The 10 most frequent species were *Ochlopoa annua* (625 occurrences), *Urtica dioica* (588), *Sonchus oleraceus* (548), *Stellaria media* (520), *Capsella bursa-pastoris* (396), *Polygonum aviculare* (373), *Dactylis glomerata* (369), *Senecio vulgaris* (336) and *Anisantha sterilis* (310). Considering only those plots with the most frequent plot size range (10-30 m^2^, n = 867 plots), the average species richness per plot was 16 (minimum = 3, maximum = 48). The class with the richest species pool was *Epilobietea angustifolii* (n = 605) and the poorest was *Bidentetea* (n = 132).

@@

The most frequent life form in the species pool (n = 1,149 species) were hemicryptophytes (44%), followed by therophytes (41%), geophytes (8%), chamaephytes (6%), phanerophytes (5%), bryophytes (1%) and hydrophytes (1%). The proportion of therophytes across classes and alliances (**Fig. 4**) agreed with the traditional description of the syntaxa: therophytes dominated the annual communities of trampled-soil vegetation (*Polygono-Poetea annuae*), crops weeds (*Papaveretea rhoeadis*, *Digitario sanguinalis-Eragrostietea minoris*) and annual ruderals (*Chenopodietea*, *Sisymbrietea*, *Bidentetea*); while perennial life forms (especially hemicryptophytes) dominated walls (*Cymbalario-Parietarietea diffusae*) and perennial ruderal vegetation (*Artemisietea vulgaris*, *Epilobietea angustifolii*). Geophytes represented a relatively high proportion of the vegetation in two alliances: *Allion triquetri* and *Senecionion fluviatilis*.

@@

The anthropogenic vegetation of the Iberian Atlantic territories is home to more than a thousand plant taxa. This is approximately one third of the whole regional species pool represented in the SIVIM database [@RN5374] and one fifth of the Iberian flora [@RN5503]. More than half of this synanthropic diversity (c. 600 taxa) occurs in mesic to wet perennial ruderal vegetation, highlighting the potential of some anthropogenic habitats as a source of biodiversity in human-dominated landscapes [@RN5446; @RN5445]. These results suggest three main objectives for the sustainable management of human-made habitats: (1) focusing efforts on alien plant management in anthropogenic plant communities of wet habitats; (2) conserving examples of archaeophyte-rich annual weed communities, especially those linked to traditional agricultural practices [@RN5502]; and (3) conserving and promoting mesic ruderal and fringe vegetation as a biodiversity refuge in urban and peri-urban areas. We hope that our synthesis of the anthropogenic communities of the Iberian Atlantic territories serves as a contribution towards a pan-European synthesis of human-made vegetation, a synthesis that could shed light on the coexistence between humans and plants.

@@

**Appendix S5** Synoptic table of the fidelity (phi) of the diagnostic species of the 25 anthropogenic alliances.

@@

```{r fig4, echo = FALSE, fig.pos = "H", fig.cap = "Proportion of therophytes, geophytes and other life forms (hemicryptophytes, chamaephytes, phanerophytes, bryophytes, hydrophytes) in the anthropogenic plant communities of the Iberian Atlantic territories. Each bar represents the plot-level average proportion of therophytes (dark shade), geophytes (medium shade) and other life forms (light shade) in each vegetation alliance. Colors indicate the vegetation class."}
knitr::include_graphics("../results/figures/F4 - lifeforms.png")
```

>>>>

* *Results* We assigned 2,086 vegetation plots to 25 anthropogenic alliances representing 9 vegetation classes (*Cymbalario-Parietarietea diffusae*, *Polygono-Poetea annuae*, *Papaveretea rhoeadis*, *Digitario sanguinalis-Eragrostietea minoris*, *Chenopodietea*, *Sisymbrietea*, *Bidentetea*, *Artemisietea vulgaris* and *Epilobietea angustifolii*). The plots included 1,149 species: 78% natives, 15% archaeophytes and 7% neophytes. Vegetation groups were organized along a principal axis of abiotic stress (dry-sunny to moist-shady habitats) and a secondary axis of disturbance.

@@

In temperate Europe, anthropogenic vegetation is a melting pot of floras from different biogeographical origins. The starting native pool of species favored by humans has been constantly enriched by human-assisted arrival of alien (i.e. non-native) species [@RN5504]. The introduction of archaeophytes (i.e. alien species arriving before 1.500 CE) likely started in the early Neolithic and peaked in the Bronze Age, with successive waves during periods of accelerated human colonization such as the early Middle Ages [@RN5504]. In Europe, synanthropic archaeophytes include many Mediterranean taxa originating from the segetal flora of the Near East [@RN5494]. Some of these archaeophytes are 'obligate weeds' that do not occur in non-anthropogenic habitats and are assumed to have evolved together with human agriculture [@RN5494]. The synanthropic flora of Europe has been further enriched by neophytes, alien taxa that arrived after the expansion of intercontinental trade in the Modern Age (after 1.500 CE) [@RN5439; @RN5451]. In the last century, the European anthropogenic vegetation has shown a decrease in species richness and diversity [@RN5423], as rare synanthropic species have become rarer and neophytes have increased their abundance at the expense of natives and archaeophytes [@RN5441].

@@

To circumscribe our study vegetation, we followed the definition of anthropogenic vegetation in the revised classification of the vegetation of Europe by Mucina *et al.* [-@RN4698]. In our study area, this potentially includes the vegetation classes *Polygono-Poetea annuae*, *Papaveretea rhoeadis*, *Digitario sanguinalis-Eragrostietea minoris*, *Chenopodietea*, *Sisymbrietea*, *Bidentetea*, *Artemisietea vulgaris* and *Epilobietea angustifolii* (which includes *Galio-Urticetea*). For the sake of completeness, we also considered the class *Cymbalario-Parietarietea diffusae* included by Mucina *et al.* [-@RN4698] in the vegetation of rock crevices and screes, since this class encompasses the vegetation of human-made walls.

@@

The main aim of our classification was for it to be consistent with regional phytosociological expertise and to match that expertise with current EuroVegChecklist alliances [@RN4698]. To achieve this, one advantage of using the SIVIM database is that a majority of the plots had been assigned to associations or alliances by the original authors [@RN5374]. However, there could be misclassification due to unclear concepts or human error. Therefore, we used TWINSPAN to revise the classification by the original authors, to define the number of clusters that better aligned with our syntaxonomical checklist and to remove outliers or misclassified plots. We started this stage of the analyses by keeping only those plots (n = 2,095) that had been assigned by the original authors of the plot to any of the 9 anthropogenic vegetation classes defined in our syntaxonomical checklist (i.e., we removed the plots without syntaxa, that had been retrieved solely by the expert system). Then, we performed a TWINSPAN classification (with the same settings as above) within each vegetation class, attempting to classify its plots into a number of clusters equal to the number of alliances of that class in our checklist. We assigned the resulting clusters to an alliance in our checklist based on their composition of diagnostic species as described in the regional literature. In some cases, this resulted in a classification into a number of clusters in which (1) more than one cluster was assigned to the same alliance in the checklist; and (2) some alliances in the checklist could not be assigned to any of the clusters (i.e. the plots that had been assigned by the original authors to that alliance were separated in different clusters).  This stage of the analyses allowed us to identify those plots in which there was an agreement between (1) our TWINSPAN-based classification (i.e. numerical classification) and (2) the classification based on the original author's syntaxon (i.e. regional phytosociological expertise). These plots (n = 1,850) were the best representatives of each alliance and became our training dataset for semi-supervised classification.

@@

Next, we conducted a semi-supervised classification [@RN5593] of the whole dataset (n = 2,508 plots) into 25 TWINSPAN-validated anthropogenic alliances. Semi-supervised classification uses a training subset of *a priori* classified vegetation plots to classify a secondary subset of unclassified plots [@RN5383]. Since our goal was to refine the classification of the whole dataset, we allowed plots from the training subset to be re-assigned to other alliances during the re-classification. In addition, semi-supervised classification can create new groups to place data points that do not match the already existing *a priori* groups, but attempts to do so resulted in new groups with no ecological significance and thus we kept the 25 alliances as final anthropogenic vegetation groups. Furthermore, we applied a noise clustering fuzzy algorithm, which allows plots to be classified either into the *a priori* alliances or into a *noise* group which includes outliers and transitional plots [@RN5382]. We set the fuzziness coefficient to a low value (m = 1) to accommodate a high number of transitional plots; and we set the distance to the noise class to d = 1. We performed this analysis with the R package *vegclust* [@RN5383]. The semi-supervised classification with noise clustering resulted in the final classification of 2,086 plots into 26 alliances and 422 plots left out in the noise group. We used pairwise PERMANOVA (with 100,000 iterations, Euclidean distances and Holm's p-value correction) fitted with the R package *RVAideMemoire* [@RN5381] to test the significance of the final vegetation alliances; along with Principal Component Analysis (PCA) as implemented in the R package *FactoMineR* [@RN3166] to visualize the relationships between the classes and alliances.

@@

We classified the species as native, archaeophytes or neophytes in the Iberian Atlantic territories. Information for archaeophyte status in the Iberian Peninsula is scarce and inconsistent. Therefore, we used the catalogues of archaeophytes for Britain [@RN5490] and the Czech Republic [@RN5626] to create a starting list of species that are considered archaeophytes in those countries. Then, we used *Flora iberica* [-@RN3243] to remove from that list those species that are considered native in the Mediterranean part of the Iberian Peninsula. We classified the remaining species as archaeophytes in the Iberian Atlantic territories. Nonetheless, we must stress that identifying archaeophytes in southern Europe is highly problematic [@RN5444] and our classification must be taken as an indication of putative archaeophyte character, solely for the purposes of vegetation comparison with other European regions, rather than a definitive classification of the species; as stated by Preston *et al.* [@RN5490], the classification of a species as an archaeophyte should be interpreted as a hypothesis to be tested by further studies.

## Species life form, height and flowering phenology

We extracted the species’ life forms from *FloraVeg.EU* [-@RN5385]. From *Flora iberica* [-@RN3243] we extracted the maximum height, median month of flowering and length of the flowering period of each species in the dataset. We used these variables to calculate plot-level community-weighted means (weighting by species cover) for each trait, to characterize the vegetation height and flowering phenology of the anthropogenic plant communities.

@@

The semi-supervised classification resulted in the classification of 2,086 vegetation plots into 25 anthropogenic alliances (**Fig. 1**, description of the alliances and translation into EUNIS habitat types in **Appendix S3**, groups of characteristic species in **Appendix S4**, synoptic table in **Appendix S5**) representing 9 vegetation classes. Of the 34 alliances in our initial syntaxonomical checklist, nine alliances were not assigned to any cluster. Of these, eight had been assigned to a few plots by the original authors: *Onopordion castellani* (n = 6 plots), *Onopordion acanthii* (5), *Euphorbion prostratae* (4), *Caucalidion lappulae* (4), *Roemerion hybridae* (2), *Hordeion murini* (2), *Taeniathero-Aegilopion geniculatae* (1) and *Fragarion vescae* (1). The alliance *Aegopodion podagrariae* had been assigned by the original authors to 39 plots, but our TWINSPAN analysis could not separate this alliance from the alliances *Geo urbani-Alliarion officinalis* and *Balloto-Conion maculati*. The regional literature included in our dataset recognized 69 associations (**Appendix S6**) within the final 25 alliances, but exploratory analysis (not shown) indicated that the separation of these associations within the alliances had a generally weak support based on numerical classification methods.

The classified plots included 1,149 species. The 10 most frequent species were *Ochlopoa annua* (625 occurrences), *Urtica dioica* (588), *Sonchus oleraceus* (548), *Stellaria media* (520), *Capsella bursa-pastoris* (396), *Polygonum aviculare* (373), *Dactylis glomerata* (369), *Senecio vulgaris* (336) and *Anisantha sterilis* (310). Considering only those plots with the most frequent plot size range (20 m^2^, n = 260 plots), the average species richness per plot was 16 (minimum = 3, maximum = 37). The class with the richest species pool was *Epilobietea angustifolii* (n = 605) and the poorest was *Bidentetea* (n = 132).

@@

The most frequent life form in the species pool (n = 1,149 species) were hemicryptophytes (44%), followed by therophytes (41%), geophytes (8%), chamaephytes (6%), phanerophytes (5%), bryophytes (1%, not consistently recorded in all plots) and hydrophytes (1%). The proportion of therophytes across classes and alliances (**Fig. 4**) agreed with the traditional description of the syntaxa: therophytes dominated the annual communities of trampled-soil vegetation (*Polygono-Poetea annuae*), crops weeds (*Papaveretea rhoeadis*, *Digitario sanguinalis-Eragrostietea minoris*) and annual ruderals (*Chenopodietea*, *Sisymbrietea*, *Bidentetea*); while perennial life forms (especially hemicryptophytes) dominated walls (*Cymbalario-Parietarietea diffusae*) and perennial ruderal vegetation (*Artemisietea vulgaris*, *Epilobietea angustifolii*). Geophytes represented a relatively high proportion of the vegetation in two alliances: *Allion triquetri* and *Senecionion fluviatilis*.

@@

The anthropogenic vegetation of the Iberian Atlantic territories is home to more than a thousand plant species. This is approximately one third of the whole regional species pool represented in the SIVIM database [@RN5374] and one fifth of the Iberian flora [@RN5503]. More than half of this synanthropic diversity (c. 600 species) occurs in mesic to wet perennial ruderal vegetation, highlighting the potential of some anthropogenic habitats as a source of biodiversity in human-dominated landscapes [@RN5446; @RN5445]. These results suggest three main objectives for the sustainable management of human-made habitats: (1) focusing efforts on alien plant management in anthropogenic plant communities of wet habitats; (2) conserving examples of archaeophyte-rich annual weed communities, especially those linked to traditional agricultural practices [@RN5502]; and (3) conserving and promoting mesic ruderal and fringe vegetation as a biodiversity refuge in urban and peri-urban areas. We hope that our synthesis of the anthropogenic communities of the Iberian Atlantic territories serves as a contribution towards a pan-European synthesis of human-made vegetation, a synthesis that could shed light on the coexistence between humans and plants.

@@

**Appendix S5** Synoptic table of the anthropogenic communities found in NW Iberia.

@@

```{r fig4, echo = FALSE, fig.pos = "H", fig.cap = "Proportion of therophytes, hemicryptophytes, geophytes and other life forms (chamaephytes, phanerophytes, bryophytes, hydrophytes) in the anthropogenic plant communities of the Iberian Atlantic territories. Each bar represents the plot-level average proportion of therophytes (dark shade),hemicryptophytes (medium-dark shade), geophytes (medium-light shade) and other life forms (light shade) in each vegetation alliance. Colors indicate the vegetation class."}
knitr::include_graphics("../results/figures/F4 - lifeforms.png")
```
